### 2021-04-30 submission/manuscript.tex
Edit results section

(Modify depending on target journal, currently abstract submitted to
World Microbe Forum) Antibiotics are a major risk factor for
\emph{Clostridioides difficile} infections (CDIs) because of their
impact on the intestinal microbiome. However, non-antibiotic medications
such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350,
also alter the microbiota, but whether PEG impacts CDI susceptibility
and clearance is unclear. To examine how PEG impacts susceptibility, we
treated C57Bl/6 mice with 5-day and 1-day doses of 15\% PEG in the
drinking water and then challenged the mice with \emph{C. difficile} 630
spores. We used clindamycin-treated mice as a control because they
consistently clear \emph{C. difficile} within 10 days post-infection
(dpi). To examine how PEG treatment impacts clearance, we administered
PEG for 1 day to clindamycin-treated, \emph{C. difficile}-challenged
mice either immediately following challenge or 3 dpi. We collected
longitudinal stool samples to examine \emph{C. difficile} levels in the
stool via anaerobic culture and profiled the microbiota by 16S rRNA
sequencing. PEG treatment alone was sufficient to render mice
susceptible to CDI and 5-day PEG-treated mice remain colonized for up to
30 dpi. Additionally, 5-day PEG treated mice remained susceptible to CDI
10-days post treatment. In contrast, 1-day PEG treated mice were
transiently colonized, clearing \emph{C. difficile} within 7 dpi.
Although 5-day PEG-treated mice exhibited prolonged \emph{C. difficile}
colonization, we saw no difference in histological inflammation between
PEG- and clindamycin-treated mice. Additionally, administering PEG to
mice after \emph{C. difficile} challenge prolonged colonization up to 30
dpi in mice that received PEG immediately after challenge and 15 dpi in
mice that received PEG 3 dpi. When we examined microbiota composition
across our different treatment groups, we found decreased richness in
the PEG-treated mice that exhibited prolonged \emph{C. difficile}
colonization. Importantly, there were increased Bacteroides and
Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in
most of the PEG-treated mice with prolonged \emph{C. difficile}
colonization. Our findings suggest the osmotic laxative PEG 3350 alters
the mouse microbiota and disrupts colonization resistance to \emph{C.
difficile}, as well as clearance in mice with a CDI. Considering that
most hospitals recommend not performing \emph{C. difficile} testing on
patients taking laxatives and laxatives are used when administering
fecal microbiota transplants via colonoscopy to patients with recurrent
CDIs, further studies are needed to evaluate if laxatives impact human
microbiota colonization resistance.

@@

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility. Our group has shown
that when \emph{C. difficile} negative controls are separated into two
groups based on stool consistency, there are shared microbiota features
such as lower alpha diversity in samples from CDI patients and control
patients with diarrhea compared to control samples that were \emph{C.
difficile} negative with non-diarrheal consistency (12). These results
led to a hypothesis that bacterial communities from patients
experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupt the human intestinal microbiota (13). The
ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} infection (14--17).
The mucus barrier is thought to mediate protection from \emph{C.
difficile} infections by protecting intestinal epithelial cells from the
toxins produced by \emph{C. difficile} (18, 19). However, whether
laxative administration results in more severe CDIs in mice and how long
mice remain colonized with \emph{C. difficile} after challenge is
unclear.

@@

In the past, our group has used C57BL6 mice to characterize how
antibiotics including clindamycin disrupt the microbiota and influence
\emph{C. difficile} susceptibility and clearance (24--26). Although, two
groups have now shown PEG treatment alone renders mice susceptible to
\emph{C. difficile} (15, 17), these studies have raised additional
questions regarding the dynamics and severity of infection as well as
the role of laxative treatment in \emph{C. difficile} clearance that
should be addressed to better inform how we think about laxatives in the
context of CDIs. Here, we used our C57BL/6 clindamycin model as a
control group to characterize how long PEG-treated mice remain
susceptible, whether PEG treatment results in more severe CDI and
sustained \emph{C. difficile} colonization, and if PEG treatment
post-CDI can promote \emph{C. difficile} clearance.

@@

\textbf{5-day laxative treatment leads to prolonged \emph{C. difficile}
colonization in mice.} We compared mice treated with the osmotic
laxative PEG 3350 to our standard 10 mg/kg clindamycin treatment, which
temporarily renders the mice susceptible to \emph{C. difficile}, with
mice typically clearing \emph{C. difficile} within 10 days
post-infection (9, 26). All PEG-treated mice were administered a 15\%
PEG solution in the drinking water for 5-days, one group was also
treated with clindamycin, and one group was allowed to recover for 10
days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss
in all 3 groups of PEG-treated mice, with the greatest change in weight
observed on the fifth day of PEG treatment and the mice recovered most
of the weight five days after treatment (Fig. 1B). After either PEG,
clindamycin, or PEG and clindamycin treatment all mice were challenged
with 10\textsuperscript{5} \emph{C. difficile} 630 spores (Fig. 1A). All
treatments rendered mice susceptible to \emph{C. difficile}
colonization. However, PEG-treated mice remained colonized with \emph{C.
difficile} at a high level through thirty days post-infection (Fig. 1C).
In contrast, the clindamycin-treated mice cleared \emph{C. difficile}
within ten days post-infection. Surprisingly, PEG-treated mice were
still susceptible to \emph{C. difficile} infection after 10-days of
recovery from treatment, although \emph{C. difficile} was not detectable
in most of the group in the initial five days post-infection (Fig. 1C).
From 8 days post-infection onward, the median \emph{C. difficile} CFU
stabilized for the 10-day recovery group of PEG-treated mice and
remained high through 30 days post-infection (Fig. 1C). Thus, osmotic
laxative treatment alone was sufficient to render mice susceptible to
prolonged \emph{C. difficile} colonization and PEG-treated mice remained
susceptible up to ten days post-treatment.

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (14--17), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time and treatment group explained half of the observed
variation between fecal communities with most of the remaining variation
explained by interactions between treatment group and other experimental
variables including time, cage effects, and sequencing preparation plate
(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
well-documented phenomenon that mice housed in the same cages have
similar microbial communities due to copraphagy (27). We tried to
minimize the impact of cage effects on our experiment by breaking up
cagemates when assigning mice to treatment groups and primarily housing
only two mice per cage. Importantly, although we conducted a total of 5
separate experiments, the experiment number and it's interaction with
treatment group was not one of the variables that significantly
explained the observed variation in fecal communities (Data Set S1,
Sheet X). Interestingly, none of the treatment groups recovered to their
baseline community structure either 10 or 30 days post-infection
suggesting other community features besides recovery to baseline were
responsible for the prolonged \emph{C. difficile} colonization in
PEG-treated mice (Fig. 2B).

Next, we examined alpha diversity by looking at Shannon diversity over
time. Although both clindamycin and PEG treatments decreased diversity,
Shannon diversity was lower in the groups of mice that received PEG
treatment compared to those that received clindamycin through thirty
days post-infection (Fig. 2C). We next examined the bacterial genera
that shifted after PEG treatment by comparing the baseline samples of
mice treated with only PEG to samples from the same mice one day
post-treatment. We found 18 genera that were altered by PEG treatment
(Data Set S1, Sheet X). The majority of the bacterial relative
abundances decreased after PEG, but \emph{Enterobacteriaceae} and
\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
unique to PEG treatment, as \emph{Bacteroides} actually decreased in
clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria
that differed across treatment groups over multiple time points. We
found 24 genera were different over multiple time points out of the 33
genera that were different between treatment groups (Fig. 2E, Data Set
S1, Sheet X). Thus, PEG had a significant impact on the fecal microbiota
that was maintained over time and was distinct from clindamycin
treatment.

Interestingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine the bacteria that changed
during the post-infection period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-infection to
detectable in the stool samples with the median stabilizing around 8
days post-infection (Fig. S1A). Interestingly, we found
\emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
\emph{Akkermansia} were changing during the time period when \emph{C.
difficile} was becoming detectable in the stools (Fig. S1B), although
none of the bacteria were significant after multiple hypothesis
correction (Data Set S1, Sheet X). While we did not identify a clear
signal to explain the delayed appearance of \emph{C. difficile} in the
5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay is striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated PEG treatment disrupts the mucus layer
and alters the immune response in mice (16), we decided to examine the
impact of PEG treatment on the mucosal microbiota and CDI severity. To
evaluate the mucosal microbiota, we sequenced snips of tissue collected
from the cecum, proximal colon, and distal colon. Similar to what was
observed with the stool samples, alpha diversity was lower in the
PEG-treated mice compared to clindamycin treatment (Fig. 3A). Although
alpha diversity continued to increase over time based on the communities
from PEG-treated mice collected at 20 and 30 days post-infection (Fig.
3A, Data Set S1, Sheet X). Group, time point, and their interactions
with other variables (cage, experiment number, and sample type)
explained the majority of the variation observed in mucosal communities
(PERMANOVA combined R\textsuperscript{2} = 0.83, \emph{P} \textless{}
0.05, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in
the mucosal microbiota between treatment groups at 6 days post-infection
with 16 genera that were significantly different in all three of the
tissue types we collected (cecum, proximal colon, and distal colon; Fig.
S3A, Data Set S1, Sheet X). However, at 30 days post-infection only
\emph{Bacteroides}, \emph{Clostridiales}, \emph{Firmicutes}, and
\emph{Ruminococcaceae} were different between treatment groups and only
in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X).
Interestingly, one of the bacteria that was different between treatment
groups at 6 days post-infection was \emph{Peptostreptococcaceae} (the
genera that contains a sequence that matches \emph{C. difficile}).
\emph{Peptostreptococcaceae} was primarily only present in the 5-day PEG
treatment group of mice and decreased in the proximal and distal colon
tissues over time (Fig. S2B). Thus, PEG treatment had a significant
impact on the mucosal microbiota and we detected \emph{C. difficile}
sequences in the cecum, proximal colon, and distal colon tissue
communities.

Next, we examined CDI severity by evaluating cecum and colon
histopathology (28) and found there was no difference in cecum and colon
scores between clindamycin and PEG-treated mice that were challenged
with \emph{C. difficile} at 4 days post-infection (Fig. 3D), the time
point typically examined in \emph{C. difficile} 630 challenged mice
(29). We also looked at 6 days post-infection because that was when we
started to see a large difference in \emph{C. difficile} colonization
levels between PEG- and clindamycin-treated mice (Fig. 1C). Although,
there was a slight difference in the colon between PEG and
clindamycin-treated mice, there was no difference in the cecum and the
overall score is still relatively low given that the max possible
summary score is twelve (Fig. 3E). Therefore, although PEG treatment had
a disruptive effect on the mucosal microbiota, the impact of \emph{C.
difficile} 630 infection on the cecum and colon was similar between PEG
and clindamycin treated mice.

\textbf{\emph{C. difficile} challenge does not have a synergistic
disruptive effect on the microbiota of PEG-treated mice} Because
\emph{C. difficile} itself can have an impact on the microbiota (30), we
also sequenced the tissue and stools of mock-challenged clindamycin and
5-day PEG treated mice. Examining the stools of the mock-challenged mice
revealed similar bacterial disruptions as the \emph{C. difficile}
challenged mice (Fig. S2A-C). Similarly, there was no difference between
the tissues of mock and \emph{C. difficile} challenged mice (Fig.
S2D-F). Thus, most of the microbiota alterations we observed in the
PEG-treated mice were a result of the laxative and not an interaction
between the laxative and \emph{C. difficile}.

\textbf{1-day laxative treatment results in transient \emph{C.
difficile} colonization and minor microbiota disruption} Next, we
decided to examine how a shorter osmotic laxative perturbation would
impact the microbiome and susceptibility to \emph{C. difficile}. We
administered either a 1-day PEG treatment, a 1-day PEG treatment with a
1-day recovery period, or clindamycin to mice before challenging them
with \emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
mice, the 1-day PEG treated mice were only transiently colonized and
cleared \emph{C. difficile} by 7 days post-infection (Fig. 3B). The
stool communities of PEG-treated mice were also only transiently
disrupted, with Shannon diversity recovering close to baseline by 7 days
post-infection (Fig. 3C-D). We found 14 bacteria were impacted by
treatment, but recovered close to baseline levels by 7 days
post-infection including \emph{Enterobacteriaceae},
\emph{Clostridiales}, \emph{Porpyromonadaceae}, and
\emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
suggest the duration of treatment matters when considering the impact of
the laxatives, with shorter treatments resulting in a transient loss of
\emph{C. difficile} colonization resistance.

\textbf{Post-CDI laxative treatment disrupts clearance in
clindamycin-treated mice regardless of whether an FMT is also
administered} Since 1-day PEG treatment resulted in a more mild
microbiota perturbation, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. We were also interested in exploring whether
PEG might help with engraftment in the context of FMTs. To examine the
impact of PEG treatment on \emph{C. difficile} clearance and FMT
treatment, we treated 4 groups of mice with clindamycin and then
challenged all mice with \emph{C. difficile} before administering the
following treatments: no additional treatment, 1-day PEG immediately
after challenge, and 1-day PEG treatment 3 days after challenge followed
by either administration of an FMT or PBS solution by oral gavage (Fig.
5A). Contrary to our hypothesis, all groups of mice that received PEG
exhibited prolonged \emph{C. difficile} colonization (Fig. 5B). Next we
examined how post-CDI PEG treatment impacted the microbiota. Alpha
diversity was lower in the PEG-treated mice with the exception of the
PEG-treated mice that were administered an FMT (Fig 5C-D). The FMT
appeared to partially restore Shannon diversity but not richness (Fig.
5C-D). Similarly, we saw some overlap between the communities of mice
that received FMT and the mice treated with only clindamycin after 5
days post-infection (Fig. 6A). The increase in Shannon diversity
suggests that the FMT did have an impact on the microbiota, despite
seeing prolonged \emph{C. difficile} colonization in the FMT treated
mice. However, only \emph{Bacteroidales} and \emph{Porphyromonadaceae}
consistently differed between the mice received either an FMT or PBS
gavage (Fig. 6B), suggesting the FMT only restored a couple of genera.
Overall, we found there were 24 bacteria that differed over multiple
time points between treatment groups including increased
\emph{Akkermansia} and decreased \emph{Ruminococcaceae},
\emph{Clostridiales}, \emph{Lachnospiraceae}, and \emph{Oscillibacter}
in mice that received PEG after \emph{C. difficile} challenge (Fig. 6C).
In sum, administering PEG actually prolonged \emph{C. difficile}
colonization, including in mice that received an FMT, which only
restored a couple of bacterial genera.

>>>>

(Modify depending on target journal, currently abstract submitted to
World Microbe Forum) Antibiotics are a major risk factor for
\emph{Clostridioides difficile} infections (CDIs) because of their
impact on the intestinal microbiome. However, non-antibiotic medications
such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350,
also alter the microbiota, but whether PEG impacts CDI susceptibility
and clearance is unclear. To examine how PEG impacts susceptibility, we
treated C57Bl/6 mice with 5-day and 1-day doses of 15\% PEG in the
drinking water and then challenged the mice with \emph{C. difficile} 630
spores. We used clindamycin-treated mice as a control because they
consistently clear \emph{C. difficile} within 10 days post-challenge
(dpc). To examine how PEG treatment impacts clearance, we administered
PEG for 1 day to clindamycin-treated, \emph{C. difficile}-challenged
mice either immediately following challenge or 3 dpc. We collected
longitudinal stool samples to examine \emph{C. difficile} levels in the
stool via anaerobic culture and profiled the microbiota by 16S rRNA
sequencing. PEG treatment alone was sufficient to render mice
susceptible to CDI and 5-day PEG-treated mice remain colonized for up to
30 dpc. Additionally, 5-day PEG treated mice remained susceptible to CDI
10-days post treatment. In contrast, 1-day PEG treated mice were
transiently colonized, clearing \emph{C. difficile} within 7 dpc.
Although 5-day PEG-treated mice exhibited prolonged \emph{C. difficile}
colonization, we saw no difference in histological inflammation between
PEG- and clindamycin-treated mice. Additionally, administering PEG to
mice after \emph{C. difficile} challenge prolonged colonization up to 30
dpc in mice that received PEG immediately after challenge and 15 dpc in
mice that received PEG 3 dpc. When we examined microbiota composition
across our different treatment groups, we found decreased richness in
the PEG-treated mice that exhibited prolonged \emph{C. difficile}
colonization. Importantly, there were increased Bacteroides and
Enterobacteriaceae and decreased Lachnospiraceae and Oscillibacter in
most of the PEG-treated mice with prolonged \emph{C. difficile}
colonization. Our findings suggest the osmotic laxative PEG 3350 alters
the mouse microbiota and disrupts colonization resistance to \emph{C.
difficile}, as well as clearance in mice with a CDI. Considering that
most hospitals recommend not performing \emph{C. difficile} testing on
patients taking laxatives and laxatives are used when administering
fecal microbiota transplants via colonoscopy to patients with recurrent
CDIs, further studies are needed to evaluate if laxatives impact human
microbiota colonization resistance.

@@

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility (10). Our group has
shown that when \emph{C. difficile} negative controls are separated into
two groups based on stool consistency, there are shared microbiota
features such as lower alpha diversity in samples from CDI patients and
control patients with diarrhea compared to control samples that were
\emph{C. difficile} negative with non-diarrheal consistency (12). These
results led to a hypothesis that bacterial communities from patients
experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupt the human intestinal microbiota (13). The
ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} infection (14--17).
The mucus barrier is thought to mediate protection from \emph{C.
difficile} infections by protecting intestinal epithelial cells from the
toxins produced by \emph{C. difficile} (18, 19). Whether laxative
administration results in more severe CDIs in mice and how long mice
remain colonized with \emph{C. difficile} after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt
the microbiota and influence \emph{C. difficile} susceptibility and
clearance (24--26). Although, two groups have now shown PEG treatment
alone renders mice susceptible to \emph{C. difficile} (15, 17), these
studies have raised additional questions regarding the dynamics and
severity of infection as well as the role of laxative treatment in
\emph{C. difficile} clearance. Addressing these questions will better
inform how we think about laxatives in the context of CDIs. Here, we
characterized how long PEG-treated mice remain susceptible, whether PEG
treatment results in more sustained \emph{C. difficile} colonization and
severe CDI and than mice treated with clindamycin, and if PEG treatment
after colonization can promote \emph{C. difficile} clearance.

@@

\textbf{5-day laxative treatment led to prolonged \emph{C. difficile}
colonization in mice.} Building off of previous work that showed
treating mice with the osmotic laxative, PEG 3350, rendered mice
susceptible to \emph{C. difficile} colonization (15, 17), we decided to
test how long \emph{C. difficile} colonization is sustained and how long
PEG-treated mice remain susceptible to \emph{C. difficile}. We compared
three groups of mice treated with PEG 3350 to one group of mice treated
with our standard 10 mg/kg clindamycin treatment, which temporarily
renders the mice susceptible to \emph{C. difficile} colonization, with
mice typically clearing \emph{C. difficile} within 10 days
post-challenge (9, 26). All three groups of PEG-treated mice were
administered a 15\% PEG solution in the drinking water for 5-days, 1.
received no additional treatment, 2. was also treated with clindamycin,
and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A).
PEG treatment resulted in weight loss in all 3 groups of PEG-treated
mice relative to their baseline weights, with the greatest change in
weight observed on the fifth day of PEG treatment. The mice recovered
most of the lost weight by five days after treatment (Fig. 1B). After
either the PEG, clindamycin, or PEG and clindamycin treatment all mice
were challenged with 10\textsuperscript{5} \emph{C. difficile} 630
spores (Fig. 1A). All treatments rendered mice susceptible to \emph{C.
difficile} colonization. However, PEG-treated mice remained colonized
with \emph{C. difficile} at a high level through thirty days
post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice
cleared \emph{C. difficile} within ten days post-challenge (Fig. 1C).
Therefore, PEG treatment led to sustained colonization in contrast to
clindamycin mice that naturally cleared \emph{C. difficile} within ten
days post-challenge.

Notably, we also found PEG-treated mice were still susceptible to
\emph{C. difficile} colonization after a 10-day recovery period,
although \emph{C. difficile} was not detectable in most of the group in
the initial five days post-challenge (Fig. 1C, S1A). One mouse was found
dead on the 6th day post-challenge, presumable due to \emph{C.
difficile} as the bacteria became detecable in that mouse on the 4th day
post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward,
the density of \emph{C. difficile} stabilized in the 10-day recovery
group and remained high through 20-30 days post-challenge (Fig. 1C).
Thus, osmotic laxative treatment alone was sufficient to render mice
susceptible to prolonged \emph{C. difficile} colonization and
PEG-treated mice remained susceptible through ten days post-treatment.

\textbf{5-day laxative treatment differentially disrupted the fecal
microbiota compared to clindamycin treatment.} Since osmotic laxatives
and clindamycin have previously been shown to disrupt the murine
microbiota (14--17), we hypothesized the different \emph{C. difficile}
colonization dynamics between mice treated with the osmotic laxative or
clindamycin were due to the two drugs having differential effects on the
microbiota. We profiled the stool microbiota over time by sequencing the
V4 region of the 16S rRNA gene to compare changes across treatment
groups. We found time and treatment group explained half of the observed
variation between fecal communities with most of the remaining variation
explained by interactions between treatment group and other experimental
variables including time, cage effects, and sequencing preparation plate
(PERMANOVA combined R\textsuperscript{2} = 0.95, \emph{P} \textless{}
0.001, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the
well-documented phenomenon that mice housed in the same cages have
similar microbial communities due to coprophagy (27). We tried to
minimize the impact of cage effects on our experiment by breaking up
cagemates when assigning mice to treatment groups. Importantly, although
we conducted a total of 5 separate experiments, the experiment number
and its interaction with treatment group was not one of the variables
that significantly explained the observed variation in fecal communities
(Data Set S1, Sheet X). None of the treatment groups recovered to their
baseline community structure either 10 or 30 days post-challenge
suggesting other community features besides recovery to baseline were
responsible for the prolonged \emph{C. difficile} colonization in
PEG-treated mice (Fig. 2B).

Since none of the communities completely recovered in the follow-up
period after treatments, we next profiled community diversity and
composition. We examined the alpha diversity dynamics by calculating the
communities' Shannon diversity. Although both clindamycin and PEG
treatments decreased diversity, Shannon diversity was lower in the
groups of mice that received PEG treatment compared to those that
received clindamycin through thirty days post-challenge (Fig. 2C). We
next examined the bacterial genera that shifted after PEG treatment by
comparing the baseline samples of mice treated with only PEG to samples
from the same mice one day post-PEG-treatment. We found 18 genera that
were altered by PEG treatment (Data Set S1, Sheet X). The majority of
the bacterial relative abundances decreased after the PEG treatment, but
the relative abundance among members of the \emph{Enterobacteriaceae}
and \emph{Bacteroides} increased. The increase in \emph{Bacteroides}
relative abundance was unique to PEG treated mice, as the
\emph{Bacteroides} relative abundance actually decreased in clindamycin
treated mice (Fig. 2D). Finally, we identified the genera whose relative
abundance differed across treatment groups over multiple time points. Of
the 33 genera that were different between treatment groups, 24 genera
were different over multiple time points (Fig. 2E, Data Set S1, Sheet
X). Thus, PEG had a significant impact on the fecal microbiota that was
maintained over time and was distinct from clindamycin treatment.

Interestingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine if there were genera that
changed during the post-challenge period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-challenge to
detectable in the stool samples with the density stabilizing around 8
days post-challenge (Fig. S1A). We found no bacteria were significantly
different over the two time periods after multiple hypothesis correction
(Data Set S1, Sheet X). However, there was also wide variation between
individual mice regarding when \emph{C. difficile} became detectable
(Fig. S1A) as well as the relative abundances of bacterial genera in the
communities (Fig. S1B). For example, two mice had a high relative
abundance of \emph{Enterobacteriaceae} throughout the post-challenge
period and this corresponded to mouse 10, which died on the sixth day
post-challenge and mouse 11, where \emph{C. difficile} was present at a
high density from the 4th day post-challenge onward (Fig. S1B). While we
did not identify a clear signal to explain the delayed appearance of
\emph{C. difficile} in the 5-day PEG mice that were allowed to recover
for 10 days prior to challenge, the delay is striking and could reflect
changes in microbial activity or metabolites that were not examined in
this study.

\textbf{5-day laxative treatment did not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated PEG treatment disrupts the mucus layer
and alters the immune response in mice (16), we decided to examine the
impact of PEG treatment on the mucosal microbiota and CDI severity. To
evaluate the mucosal microbiota, we sequenced snips of tissue collected
from the cecum, proximal colon, and distal colon. Similar to what was
observed with the stool samples, alpha diversity was lower in the
PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha
diversity continued to increase over time with the PEG-treated mice
collected at 20 and 30 days post-challenge (Fig. 3A, Data Set S1, Sheet
X). Group, time point, and their interactions with other variables
(cage, experiment number, and sample type) explained the majority of the
variation observed in mucosal communities (PERMANOVA combined
R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig. 3B, Data
Set S1, Sheet X). We saw the greatest difference in the relative
abundance of the mucosal microbiota between treatment groups
(clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days
post-challenge with 10 genera that were significantly different
(\emph{P} `r format.pval(tissue\_genera\_max\_p, eps = 0.05)) in all
three of the tissue types we collected (cecum, proximal colon, and
distal colon; Fig. S2A, Data Set S1, Sheet X). Interestingly,
\emph{Peptostreptococcaceae} (the family with a sequence that matches
\emph{C. difficile}) was one of the genera that had a significant
difference in relative abundance between treatment groups at 6 days
post-challenge. This population was primarily only present in the 5-day
PEG treatment group of mice and decreased in the proximal and distal
colon tissues over time (Fig. S2B). By 30 days post-challenge, only the
relative abundances of \emph{Bacteroides}, \emph{Clostridiales},
\emph{Firmicutes}, and \emph{Ruminococcaceae} were different between
treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data
Set S1, Sheet X). Thus, PEG treatment had a significant impact on the
mucosal microbiota and we detected \emph{C. difficile} sequences in the
cecum, proximal colon, and distal colon tissue communities.

Next, we examined the severity of \emph{C. difficile} challenge by
evaluating cecum and colon H\&E stained histopathology (28) and found
there was no difference in cecum and colon scores between clindamycin
and PEG-treated mice that were challenged with \emph{C. difficile} at 4
days post-challenge (Fig. 3D), the time point typically examined in
\emph{C. difficile} 630 challenged mice (29). We also looked at 6 days
post-challenge because that was when there was a large difference in
\emph{C. difficile} density between PEG- and clindamycin-treated mice
(Fig. 1C). Although, there was a slight difference in the colon between
PEG and clindamycin-treated mice, there was not a signifant difference
in the cecum and the overall score was relatively low (1.5-2.5 out of
12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect
on the mucosal microbiota, the impact of \emph{C. difficile} 630
challenge on the cecum and colon was similar between PEG and clindamycin
treated mice.

\textbf{\emph{C. difficile} challenge did not have a synergistic
disruptive effect on the microbiota of PEG-treated mice.} Because
\emph{C. difficile} itself can have an impact on the microbiota (30), we
also sequenced the tissue and stools of mock-challenged clindamycin and
5-day PEG treated mice. Examining the stools of the mock-challenged mice
revealed similar bacterial disruptions as the \emph{C. difficile}
challenged mice (Fig. S3A-C). Similarly, there was no difference between
the tissues of mock and \emph{C. difficile} challenged mice (Fig.
S3D-F). Thus, most of the microbiota alterations we observed in the
PEG-treated mice were a result of the laxative and not an interaction
between the laxative and \emph{C. difficile}.

\textbf{1-day laxative treatment resulted in transient \emph{C.
difficile} colonization and minor microbiota disruption.} Next, we
examined how a shorter osmotic laxative perturbation would impact the
microbiome and susceptibility to \emph{C. difficile}. We administered
either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day
recovery period, or clindamycin to mice before challenging them with
\emph{C. difficile} (Fig. 3A). In contrast to the 5-day PEG treated
mice, the 1-day PEG treated mice were only transiently colonized and
cleared \emph{C. difficile} by 7 days post-challenge (Fig. 3B). The
stool communities of PEG-treated mice were also only transiently
disrupted, with Shannon diversity recovering by 7 days post-challenge
(Fig. 3C-D). We found the relative abundances of 14 genera were impacted
by treatment, but recovered close to baseline levels by 7 days
post-challenge including \emph{Enterobacteriaceae},
\emph{Clostridiales}, \emph{Porpyromonadaceae}, and
\emph{Ruminococcaceae} (Fig. 3E, Data Set S1, Sheet X). These findings
suggest the duration of the PEG treatment was relevant, with shorter
treatments resulting in a transient loss of \emph{C. difficile}
colonization resistance.

\textbf{Post-CDI laxative treatment disrupted clearance in
clindamycin-treated mice regardless of whether an FMT was also
administered.} Since a 1-day PEG treatment resulted in a more mild
microbiota perturbation, we decided to use the 1-day treatment to
examine the hypothesis that PEG helps to flush \emph{C. difficile}
spores from the intestine. To examine the impact of PEG treatment on
\emph{C. difficile} clearance, we treated 4 groups of mice with
clindamycin and then challenged all mice with \emph{C. difficile} before
administering the following treatments: no additional treatment, 1-day
PEG immediately after challenge, and 1-day PEG treatment 3 days after
challenge followed by either administration of an FMT or PBS solution by
oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice
that received PEG exhibited prolonged \emph{C. difficile} colonization
(Fig. 5B). Next we examined how post-CDI PEG treatment impacted
microbiota diversity. Alpha diversity was lower in the PEG-treated mice
with the exception of the PEG-treated mice that were administered an FMT
(Fig 5C-D).

We were also interested in exploring whether PEG might help with
engraftment in the context of FMTs. The FMT appeared to partially
restore Shannon diversity but not richness (Fig. 5C-D). Similarly, we
saw some overlap between the communities of mice that received FMT and
the mice treated with only clindamycin after 5 days post-challenge (Fig.
6A). The increase in Shannon diversity suggests that the FMT did have an
impact on the microbiota, despite seeing prolonged \emph{C. difficile}
colonization in the FMT treated mice. However, only the relative
abundances of \emph{Bacteroidales} and \emph{Porphyromonadaceae}
consistently differed between the mice received either an FMT or PBS
gavage (Fig. 6B), suggesting the FMT only restored a couple of genera.
Overall, we found the relative abundances of 24 genera were different
between treatment groups over multiple timepoints. For example, the
relative abundance of \emph{Akkermansia} was increased and the relative
abundances of \emph{Ruminococcaceae}, \emph{Clostridiales},
\emph{Lachnospiraceae}, and \emph{Oscillibacter} were decreased in mice
that received PEG after \emph{C. difficile} challenge relative to
clindamycin treated mice (Fig. 6C). In sum, administering PEG actually
prolonged \emph{C. difficile} colonization, including in mice that
received an FMT, which only restored 2 bacterial genera.
